---
title: "211-Discussion II"
author: "Yu-Shiuan (Lily) Huang"
date: "Fall 2023"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      tidy.opts=list(width.cutoff = 80),
                      tidy = TRUE)
options(width = 80)

colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

For today's discussion, we'll first briefly go over the solution keys to Problem Set I and then continue our materials for exploring a new dataset using R.

#### 1. Solution Keys to Problem Set I

##### Question 1 (4 points)

Load the `dbj.csv` file.

-   What is the unit of observation in the data set?

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("The unit of analysis in the data set is per judge.", "darkblue")`

    You can also double check this answer by using the `unique()` on the `name` variable to see if the number of observations is equal to the number of categories in that variable.

    ```{r, message=F, warning=F, error=F}
    library(tidyverse)

    dbj <- read_csv("/Users/yu-shiuanhuang/Desktop/method-sequence/data/dbj.csv")  ## you can also use read.csv

    length(unique(dbj$name)) == nrow(dbj)
    ```

    Doing this helps us confirm that this dataset is structured around the information of individual judges. However, it's worth considering that the dataset might be organized by judicial cases, where one judge could be associated with multiple cases. In such a scenario, the unit of analysis wouldn't be per judge but per case.

-   How many judges are there in the data set, and how many variables does the data set have?

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("There are 224 judges and 12 variables in the data set.", "darkblue")`

    ```{r, message=F, warning=F, error=F}
    ## Number of judges
    nrow(dbj) ## this is just the number of observations

    ## Number of variables
    ncol(dbj)
    ```

-   How many judges in the data set have children? To answer this, first create a new binary (or "dummy") variable that is equal to 1 if a judge has any children, and 0 otherwise.

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("Out of the total of 244 judges, 199 have children.", "darkblue")`

    ```{r, message=F, warning=F, error=F}
    ## Judges with children (using a new dummy variable)
    dbj <- dbj %>% 
      mutate(has_children = ifelse(child > 0, 1, 0)) 

    dbj_c <- dbj %>%
      count(has_children)

    knitr::kable(dbj_c, "simple", align = "cc", 
             col.names = c("Has Children or Not", "Number of Judges"))
    ```

-   Of the judges that have children, what percentage has female children? To calculate this number, first create a new binary (or "dummy") variable that is equal to 1 if a judge has female children, and 0 otherwise.

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("Among the judges who have children, approximately 81.41% of them have female children.", "darkblue")`

    ```{r, message=F, warning=F, error=F}
    ## Share of judges with children that have female children
    ## First, create dummy
    dbj <- dbj %>% 
      mutate(has_female_children = ifelse(girls > 0, 1, 0))

    ## We need the average of "has_female_children" conditional on "has_children" equal to 1
    dbj_with_children <- dbj %>% 
      filter(has_children == 1)

    mean(dbj_with_children$has_female_children)
    ```

##### Question 2 (4 points)

Our outcome in this exercise will be the proportion of pro-feminist decisions, `progressive.vote`.

First, calculate the mean, median and standard deviation of this variable (across all judges).

`r colorize("**Answer:**", "darkblue")`

```{r, message=F, warning=F, error=F}
## Summary stats for progressive.vote
mean(dbj$progressive.vote)
median(dbj$progressive.vote)
sd(dbj$progressive.vote)
```

Next, we want to assess differences between groups. Create a new factor variable that takes on separate values for each of the four groups (Republican men/women, Democratic men/women) defined by gender and partisanship. The variable should be a "factor" or "character" type variable in R - you can use string labels such as `Dem_Male`, `Dem_Female`, etc. It should **not** be a numeric variable.

Having created this variable, now create a **boxplot** that summarizes the propensity to decide in a pro-feminist way by group, in one plot. Briefly discuss the results -- what does the boxplot show, and what does this tell us about differences between groups? Are partisan differences larger than gender differences, or the other way around? A brief answer is sufficient here.

`r colorize("**Answer:**", "darkblue")`

`r colorize("The plot shows that progressive decisions are more prevalent among Democratic judges than among Republican judges. Gender differences are less pronounced than partisan differences, since within parties, male and female judges have similar average proportions of pro-feminist decisions.", "darkblue")`

```{r, message=F, warning=F, error=F, fig.align='center', fig.width=6, fig.height=4}
## Create new variable
dbj <- dbj %>% 
  mutate(group = case_when(republican == 1 & woman == 1 ~ "Rep_Female",
                           republican == 0 & woman == 1 ~ "Dem_Female",
                           republican == 1 & woman == 0 ~ "Rep_Male",
                           republican == 0 & woman == 0 ~ "Dem_Male"))

## Use ggplot to plot the average by group
ggplot(dbj, aes(x = group, y = progressive.vote)) +
  geom_boxplot() +
  theme_bw() 
```

You can enhance the boxplot by including additional information, such as displaying the median of the progressive vote for each group.

```{r, message=F, warning=F, error=F, fig.align='center', fig.width=6, fig.height=4}
## Calculate median for each group
dbj_md <- dbj %>%
  group_by(group) %>%
  summarise(MD = median(progressive.vote)) 

## Use ggplot to plot the average by group with median for each group
ggplot(data = dbj, aes(x = group, y = progressive.vote)) +
  geom_boxplot() +
  geom_text(data = dbj_md, aes(group, MD, label = round(MD, 4)),
            position = position_dodge(width = 0.8), 
            size = 3, vjust = -0.5) +
  theme_bw() 
```

##### Question 3 (2 points)

From before, we have a new binary variable which takes a value of `1` if a judge has *at least* one child (that is, any children at all), and `0` otherwise.

Use this variable to answer the following questions.

-   Are Republicans and Democrats equally likely to be parents (that is, have at least one child)?

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("Republicans and Democrats are almost equally likely to be parents, with 88% of Republicans and 89% of Democrats having children.", "darkblue")`

    ```{r, message=F, warning=F, error=F}
    ## First question: average likelihood of having kids among D and R
    df <- dbj %>% 
      group_by(republican) %>% 
      summarise(mean(has_children))

    knitr::kable(df, "simple", align = "cc", 
             col.names = c("Republican", 
                           "Average Likelihood of Having Kids"))
    ```

-   Do judges with children vote differently on feminist issues than judges without? If so, how are they different?

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("It does not appear as if there are large differences in pro-feminist voting when comparing judges with and without children. Among judges without children, 46% voted in favor of pro-feminist issues, slightly higher than the 43% of judges with children who voted for pro-feminist issues.", "darkblue")`

    ```{r, message=F, warning=F, error=F}
    ## Next question: voting behavior conditional on having children
    df <- dbj %>% 
      group_by(has_children) %>% 
      summarise(mean(progressive.vote))

    knitr::kable(df, "simple", align = "cc", 
             col.names = c("Has Children", 
                           "Average Likelihood of Voting Pro-feminist Issues"))
    ```

    ```{r, message=F, warning=F, error=F, fig.align='center', fig.width=6, fig.height=4}
    ## We can also again use a boxplot for this

    ## Calculate median for each group 
    dbj_hc_md <- dbj %>%
      group_by(has_children) %>%
      summarise(MD = median(progressive.vote)) 

    ggplot(dbj, aes(factor(has_children), progressive.vote)) +
      geom_boxplot() +
      geom_text(data = dbj_hc_md, aes(factor(has_children), MD, 
                                  label = round(MD, 4)),
            position = position_dodge(width = 0.8), 
            size = 3, vjust = -0.5) +
      xlab("Judge has any children (1 = yes)") +
      ylab("Progressive decisions") +
      theme_bw()
    ```

##### Question 4 (2 points)

Please evaluate the following statements. For each statement, indicate whether you believe the statement makes sense or not. If it does not, please provide a short motivation.

1.  The judge's religion can be described as an ordinal variable.

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("Incorrect: higher values do not indicator higher ranks, since the variable does not measure a concept that is ordered.", "darkblue")`

2.  A colleague argues that the judge's religion may be an important predictor of judges' decisions -- more religious judges may decide differently than less religious judges. Therefore, your colleague recommends splitting the `religion` variable (see definition above) into two groups: the set of judges for which `religion` is greater than 10, and the set of judges for which it is 10 or lower. Then, your colleague recommends comparing the mean proportion of pro-feminist decisions in each of the two groups.

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("This does not make sense - splitting the variable in this way tells us nothing about the whether a judge is more or less religious. The variable (although it is coded as a numeric variable in the data set) is actually categorical, and only tells us which religious denomination a judge belongs to. There is actually no variable in the data that measure whether judges are more or less religious.", "darkblue")`

3.  A colleague comments that the data set is ambiguous, since there is no variable that measures whether a judge was appointed by a Democratic president.

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("This does not make sense. Since judges can only be appointed by Republican or Democratic presidents, having a single binary variable", "darkblue")` (`republican`) `r colorize("is sufficient.", "darkblue")`

##### Additional question (required for students in the methods subfield)

##### Question 5 (3 points)

What is the difference in the proportion of pro-feminist decisions between judges who have at least one daughter and those who do not have any? Compute this difference in two ways; (1) using any judge who has children, (2) separately for judges that have one, two, or three children. For (2), you should end with three estimates: one for the judges with one child, one for the judges with two children, and one for the judges with three children (HINT: you will subset the data quite a few times to achieve this).

-   Compute the difference (1) using any judge who has children.

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("The difference in the proportion of pro-feminist decisions between judges who have at least one daughter and those who do not have any is about 11%.", "darkblue")`

    ```{r, message=F, warning=F, error=F}
    dbj %>% 
      filter(has_children == 1) %>% 
      group_by(has_female_children) %>% 
      summarise(mean_prog = mean(progressive.vote)) %>% 
      summarise(diff = mean_prog[has_female_children == 1] - 
              mean_prog[has_female_children == 0])
    ```

-   Compute the difference (2) separately for judges that have one, two, or three children.

    `r colorize("**Answer:**", "darkblue")`

    `r colorize("Among judges who have at least one child, two children, and three children, the difference in the proportion of pro-feminist decisions between those who have at least one daughter and those who do not have any is about 24%, 8%, and 19%, respectively.", "darkblue")`

    ```{r, message=F, warning=F, error=F}
    diff_list <- dbj %>% 
      filter(child %in% c(1,2,3)) %>%  ## Subset to people w/ 1,2,3 children
      group_by(has_female_children, child) %>%  
      summarise(mean_prog = mean(progressive.vote)) %>%  ## Average
      ungroup() %>% 
      group_by(child) %>%  ## Calculate differences by no. of children
      summarise(diff = mean_prog[has_female_children == 1] - 
              mean_prog[has_female_children == 0])

    knitr::kable(diff_list, "simple", align = "cc", 
             col.names = c("Has no. Children", "Difference"))
    ```

#### 2. Exploring a New Dataset using R (Part II)

For today, we are using `USDC-DATASET-JOP.csv`. You can download this csv file from Canvas or from [here](https://dataverse.harvard.edu/file.xhtml?fileId=4306747&version=1.0.)

```{r, message=F, warning=F, error=F}
## Load packages
library(tidyverse)

## Import data
df <- read_csv("/Users/yu-shiuanhuang/Desktop/method-sequence/data/USDC-DATASET-JOP.csv") 
```

##### a. Summary Statistics

In the following, we will compute summary statistics for the univariate distribution of the variable that quantifies the number of defendants named in each court case (it's named `def_count`).

1.  Calculate the mean

    ```{r, message=F, warning=F, error=F}
    ## This is the easy way to calculate the mean:
    mean(df$def_count, na.rm = TRUE)
    ```

    ```{r}
    ## How to calculate it in the long/manual way?




    ## Try out!
    ```

    Recall the sample mean equation from the lecturer slide:

    $$\bar{x} = \frac{1}{N}\sum_{i=1}^{N}x_{i}=\frac{x_{1}+x_{2}+...+x_{N}}{N}$$

2.  Calculate the median

    ```{r, message=F, warning=F, error=F}
    ## This is the easy way to calculate the median:
    median(df$def_count, na.rm = TRUE)
    ```

    ```{r}
    ## How to calculate it in the long/manual way?
    ## First, sort the distribution from smallest to largest values



    ## Next, identify the "middle" value, which is (N+1)/2  



    ## Try out!
    ```

    Since the distribution of `def_count` has an odd number, we have one median. If instead we had an even number, notice that R will automatically take **the average of the two middle values**. Technically, as Gailmard (2014) points out, any number weakly between the two middle values is a median.

    ```{r, message=F, warning=F, error=F}
    fake.data <- c(3, 0, 2, 10, 7, 1) ## notice: even number of values!

    sort(fake.data)
    median(fake.data) 
    ```

    R tells us the median is 2.5 but technically it's any number on interval $[2, 3]$.

3.  Calculate the variance and standard deviation

    ```{r, message=F, warning=F, error=F}
    ## This is the easy way to calculate the variance and std. dev.:
    var(df$def_count, na.rm = TRUE)
    sd(df$def_count, na.rm = TRUE)
    ```

    ```{r}
    ## How to calculate them in the long/manual way?
    ## Sample variance



    ## Sample standard deviation



    ## Try out!
    ```

    Recall the sample variance equation from the lecturer slide:

    $$s^2_{x}=\frac{1}{N-1}\sum_{i=1}^{N}(x_i-\bar{x})^2$$

Now that you know how to manually compute the sample mean, median, variance, and standard deviation in R, let's create a custom `function` that can calculate these four descriptive statistics all at once, without relying on the built-in R functions like `mean()`, `median()`, `var()`, or `sd()`.

Please manually create a function named `des_stat` by using `function()`:

```{r}
## Create your own function

















## Try out!
```

4.  Calculate the interquartile range (IQR)

    The $n$th percentile is the value in a distribution where $n$ percent of the values are equal to or below that value. IQR is defined by the difference between the first and the third quartile.

    $$ IQR = P^{75} - P^{25} $$

    ```{r, message=F, warning=F, error=F}
    ## This is the easy way to find the first and the third quartile:
    quantile(df$def_count, 0.25) ## First quartile
    quantile(df$def_count, 0.75) ## Third quartile
    ```

    ```{r, message=F, warning=F, error=F}
    ## You can also use quantile function to find an arbitrary percentile
    quantile(df$def_count, 0.87) ## the 87th percentile
    ```

    ```{r, message=F, warning=F, error=F}
    ## Notice: R names the values it returns. That's helpful, but sometimes annoying.
    ## Remove the name this way:
    unname(quantile(df$def_count, 0.25))
    unname(quantile(df$def_count, 0.75))
    ```

    ```{r, message=F, warning=F, error=F}
    ## This is the easy way to calculate IQR:
    IQR(df$def_count)

    ## This is a longer way:
    unname(quantile(df$def_count, 0.75)) - unname(quantile(df$def_count, 0.25))
    ```

    ```{r}
    ## How to calculate IQR in the long/manual way (similar to the median)?

    ## First, sort the distribution from smallest to largest values



    ## Next, identify the value at the 25th percentile: 0.25 * N (round this up to nearest integer using ceiling function)



    ## Try out!
    ```

5.  Find outliers

    An arbitrary (but widely used) definition of an outlier is that any value in a distribution that is at least $1.5×IQR$ above the third quartile or below the first quartile.

    How many outliers are there in the distribution of `def_count`?

    ```{r, message=F, warning=F, error=F}
## First, what is the lowest and highest threshold for outliers?
iqr.d <- IQR(df$def_count)
min.threshold <- quantile(df$def_count, 0.25) - (1.5 * iqr.d)
min.threshold ## -2.5
max.threshold <- quantile(df$def_count, 0.75) + (1.5 * iqr.d)
max.threshold ## 9.5
    ```

    Any data point that is smaller than -2.5 or larger than 9.5 is an outlier.

    ```{r}
## Please count how many outliers are there in the distribution of def_count?




## Try out!
    ```


##### b. Plot distribution (histograms)

To better observe the distribution of a variable, plot histograms! Histograms visualize quantitative data or numerical data (cardinal variables).

```{r, message=F, warning=F, error=F}
## First, we would like to define the bins that will be used later
my.min.value <- min(df$def_count) ## Find the minimum of the variable
my.max.value <- max(df$def_count) ## Find the maximum of the variable

my.bins <- seq(my.min.value, my.max.value, 1) 
## Generate a sequence of number that the values are from the minimum to the maximum while the increment of the sequence is 1
```

```{r, message=F, warning=F, error=F, fig.align='center', fig.width=6, fig.height=4}
# Now, let's build a plot layer by layer:
ggplot(df) + 
  geom_histogram(aes(x = def_count), breaks = my.bins)
```

This histogram sucks! Why? Because it's not easy to see the most important part of the distribution (the low values). In other words: we're plotting a lot of outliers.

Let's try to focus in on the part of the plot that's the most interesting. To do this, we need to exclude outliers from the data and only plot any values that are not outliers!

```{r, message=F, warning=F, error=F}
## Remember: we already figured out what the outliers are:
outliers <- df$def_count[df$def_count < min.threshold | df$def_count > max.threshold]
unique(sort(outliers))
```

Looks like any number 10 or above is an outlier.

```{r, message=F, warning=F, error=F, fig.align='center', fig.width=6, fig.height=4}
## So, let's tell R to only plot a histogram for values 9 and lower.
my.max.value <- 9

## Let's make the x axis numbers correspond to the bins. 
## To do this, let's define what we want our labels to be:
my.labs <- seq(my.min.value, my.max.value, 1)
```

```{r, message=F, warning=F, error=F, fig.align='center', fig.width=6, fig.height=4}
ggplot2::ggplot(data = df, aes(x = def_count)) + 
  geom_histogram(breaks = my.bins, color = "black", fill = "gray85") + 
  scale_x_continuous(limits = c(my.min.value, my.max.value), breaks = my.labs) +
  scale_y_continuous(breaks = seq(0,30000,2500), limits = c(0,30000)) +
  labs(x = "Number of Defendants In Each Case", y = "Frequency") + 
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```

When using `ggplot2` to plot histograms, the frequency of each bar is calculated as only including the upper but not the lower bound. For example, for the bar 0-1, the frequency is $0<$`def_count`$<=1$.

We can also manually check the exact frequency of each bar to see if 0 is excluded from the 0-1 bar.

```{r, message=F, warning=F, error=F}
df %>% count(def_count)
```

We can also add vertical lines that specify mean and median of the variable.

```{r, message=F, warning=F, error=F, fig.align='center', fig.width=6, fig.height=4}
# Let's save the mean and median as objects that we can use later.
my.mean <- mean(df$def_count)
my.median <- median(df$def_count)

ggplot(df, aes(x = def_count)) + 
  geom_histogram(breaks = my.bins, color = "black", fill = "gray85") + 
  scale_x_continuous(limits = c(my.min.value, my.max.value), breaks = my.labs) + 
  scale_y_continuous(breaks = seq(0,30000,2500), limits = c(0,30000)) + 
  geom_vline(aes(xintercept = my.mean), size = 0.8, linetype = 2, color = "blue") +  
  ## add the line of mean   
  geom_vline(aes(xintercept = my.median), size = 0.8, linetype = 2, color = "red") + 
  ## add the line of median
  labs(x = "Number of Defendants In Each Case", y = "Frequency") + 
  theme_bw() + 
  theme(panel.grid.minor = element_blank())
```

We can see that the distribution of `def_count` is **right skewed**, which is also referred to positively skewed. What does this mean?

A skewed distribution occurs when one tail is longer than the other. Skewness defines the asymmetry of a distribution. Unlike the familiar normal distribution with its bell-shaped curve, these distributions are asymmetric. The two halves of the distribution are not mirror images because the data are not distributed equally on both sides of the distribution's peak.

How to Tell if a Distribution is Left Skewed or Right Skewed?

-   **Right skewed** distributions occur when the long tail is on the right side of the distribution. Analysts also refer to them as positively skewed. This condition occurs because probabilities taper off more slowly for higher values. Consequently, you'll find extreme values far from the peak on the high end more frequently than on the low. When the distribution of data is skewed to the right, *the mean is often larger than the median*.

-   **Left skewed** distributions occur when the long tail is on the left side of the distribution. Statisticians also refer to them as negatively skewed. This condition occurs because probabilities taper off more slowly for lower values. Therefore, you'll find extreme values far from the peak on the low side more frequently than the high side. When the distribution of data is skewed to the right, *the median is often greater than the mean*.

How you plot frequency is important!

What if I wanted to use different bin sizes?

```{r, message=F, warning=F, error=F, fig.align='center', fig.width=6, fig.height=4}
my.bins2 <- c(0,1,2.5,3.5,10)

ggplot(df, aes(x = def_count)) + 
  geom_histogram(breaks = my.bins2, color = "black", fill = "gray85") + 
  scale_x_continuous(breaks = my.labs) + 
  scale_y_continuous(breaks = seq(0,30000,2500), limits = c(0,30000)) +
  labs(x = "Number of Defendants In Each Case", y = "Frequency") + 
  theme_bw() + 
  theme(panel.grid.minor = element_blank())
```

This feels like it's misleading, because it is!

Let's fix it by using density on the y axis. Density is relative frequency per unit on the x axis. The density of each bin is calculated by dividing its relative frequency by its width in terms of units of the x variable (i.e., you divide the frequency of a group by the width of it).

```{r, message=F, warning=F, error=F, fig.align='center', fig.width=6, fig.height=4}
ggplot(df) + 
  geom_histogram(aes(y = ..density.., x = def_count), breaks = my.bins2,
                 color = "black", fill = "gray85") + 
  scale_x_continuous(breaks = my.labs) + 
  scale_y_continuous(breaks = seq(0,0.3,0.05), limits = c(0,0.3)) +
  labs(x = "Number of Defendants In Each Case", y = "Density") + 
  theme_bw() + 
  theme(panel.grid.minor = element_blank())
```
