---
title: "Problem Set 1"
author: "Pol211, UC Davis"
date: "Due by 10/9/23, 2023 at 11:59 PM"
output: 
  pdf_document: default
---

In this problem set, we will analyze the relationship between various demographic traits and pro-feminist voting behavior among circuit court judges. In a paper, Adam N. Glynn and Maya Sen argue that having a female child causes circuit court judges to make more pro-feminist decisions. The paper can be found at:

> Glynn, Adam N., and Maya Sen. (2015). ["Identifying Judicial Empathy: Does Having Daughters Cause Judges to Rule for Women's Issues?."](https://doi.org/10.1111/ajps.12118) *American Journal of Political Science* Vol. 59, No. 1, pp. 37--54.

The dataset `dbj.csv` contains the following variables about individual judges:

| Name               | Description                                                                                                                                                                                                                                                                                        |
|-----------------|-------------------------------------------------------|
| `name`             | The judge's name                                                                                                                                                                                                                                                                                   |
| `child`            | The number of children each judge has.                                                                                                                                                                                                                                                             |
| `circuit.1`        | Which federal circuit the judge serves in.                                                                                                                                                                                                                                                         |
| `girls`            | The number of female children the judge has.                                                                                                                                                                                                                                                       |
| `progressive.vote` | The proportion of the judge's votes on women's issues which were decided in a pro-feminist direction.                                                                                                                                                                                              |
| `race`             | The judge's race (`1` = white, `2` = African-American, `3` = Hispanic, `4` = Asian-American).                                                                                                                                                                                                      |
| `religion`         | The judge's religion (`1` = Unitarian, `2` = Episcopalian, `3` = Baptist, `4` = Catholic, `5` = Jewish, `7` = Presbyterian, `8` = Protestant, `9` = Congregationalist, `10` = Methodist, `11` = Church of Christ, `16` = Baha'i, `17` = Mormon, `21` = Anglican, `24` = Lutheran, `99` = unknown). |
| `republican`       | Takes a value of `1` if the judge was appointed by a Republican president, `0` otherwise. Used as a proxy for the judge's party.                                                                                                                                                                   |
| `sons`             | The number of male children the judge has.                                                                                                                                                                                                                                                         |
| `woman`            | Takes a value of `1` if the judge is a woman, `0` otherwise.                                                                                                                                                                                                                                       |
| `X`                | Indicator for the observation number.                                                                                                                                                                                                                                                              |
| `yearb`            | The year the judge was born.                                                                                                                                                                                                                                                                       |

\normalsize

**Note:** if you are asked to provide an interpretation or an explanation, 1-3 sentences are sufficient. There is no need to write more than that. Short, succinct answer are preferred.

*Note*

\newpage

## Question 1 (4 points)

Load the `dbj.csv` file.

-   What is the unit of observation in the data set?
-   How many judges are there in the data set, and how many variables does the data set have?
-   How many judges in the data set have children? To answer this, first create a new binary (or "dummy") variable that is equal to 1 if a judge has any children, and 0 otherwise.
-   Of the judges that have children, what percentage has female children? To calculate this number, first create a new binary (or "dummy") variable that is equal to 1 if a judge has female children, and 0 otherwise.


```{r, message=F, warning=F, error=F}


```

## Question 2 (4 points)

Our outcome in this exercise will be the proportion of pro-feminist decisions, `progressive.vote`.

First, calculate the mean, median and standard deviation of this variable (across all judges)

Next, we want to assess differences between groups. Create a new factor variable that takes on separate values for each of the four groups (Republican men/women, Democratic men/women) defined by gender and partisanship. The variable should be a "factor" or "character" type variable in R - you can use string labels such as `Dem_Male`, `Dem_Female`, etc. It should **not** be a numeric variable.

Having created this variable, now create a **boxplot** that summarizes the propensity to decide in a pro-feminist way by group, in one plot. Briefly discuss the results -- what does the boxplot show, and what does this tell us about differences between groups? Are partisan differences larger than gender differences, or the other way around? A brief answer is sufficient here.

```{r, message=F, warning=F, error=F}


```

## Question 3 (2 points)

From before, we have a new binary variable which takes a value of `1` if a judge has *at least* one child (that is, any children at all), and `0` otherwise.

$2^3$

Use this variable to answer the following questions.

-   Are Republicans and Democrats equally likely to be parents (that is, have at least one child)?
-   Do judges with children vote differently on feminist issues than judges without? If so, how are they different?

```{r, message=F, warning=F, error=F}



```

## Question 4 (2 points)

Please evaluate the following statements. For each statement, indicate whether you believe the statement makes sense or not. If it does not, please provide a short motivation.

1.  The judge's religion can be described as an ordinal variable.

2.  A colleague argues that the judge's religion may be an important predictor of judges' decisions -- more religious judges may decide differently than less religious judges. Therefore, your colleague recommends splitting the `religion` variable (see definition above) into two groups: the set of judges for which `religion` is greater than 10, and the set of judges for which it is 10 or lower. Then, your colleague recommends comparing the mean proportion of pro-feminist decisions in each of the two groups.

3.  A colleague comments that the data set is ambiguous, since there is no variable that measures whether a judge was appointed by a Democratic president.

# Additional question (required for students in the methods subfield)

## Question 5 (3 points)

What is the difference in the proportion of pro-feminist decisions between judges who have at least one daughter and those who do not have any? Compute this difference in two ways; (1) using any judge who has children, (2) separately for judges that one, two, or three children. For (2), you should end with three estimates: one for the judges with one child, one for the judges with two children, and one for the judges with three children (HINT: you will subset the data quite a few times to achieve this).

```{r}

```
